\documentclass{article}
\usepackage{graphicx}

\usepackage{amsmath}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{listings}
\lstnewenvironment{code}{\lstset{language=Haskell,basicstyle=\small\ttfamily}}{}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}

\usepackage{array}
% This is needed because raggedright in table elements redefines \\:
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\let\PBS=\PreserveBackslash

\setcounter{secnumdepth}{0}


\begin{document}
\noindent
\begin{tabular}{lr}
CHALMERS TEKNISKA H\"OGSKOLA &Tuesday ?th August, 2010.\\
Dept. of Computer Science and Engineering & Programming Paradigms\\
John Hughes                  & DAT120(CTH) / DIT330(GU) \\
\end{tabular}

\vspace{2.5cm} \noindent
\begin{center} {\LARGE
Exam in Programming Paradigms}
\end{center}

\vspace{1.5cm}

\noindent
Tuesday ??th August, 2010, EM.\\
Lecturer: John Hughes, tel 070 756 3760.
\vspace{1cm}

\noindent
Permitted aids:\\
English-Swedish or English-other language dictionary.

There are five questions, one on each paradigm, worth 12 points each
for a total of 60 points. 24 points is required to pass (grade 3), 36
points is required for grade 4, and 48 points is required for grade 5.

%\newcommand{\comment}[1]{}
\newcommand{\comment}[1]{\marginpar{#1}}

\newpage

\section{Imperative Programming [12 points]}

In this part of the exam, we use the following notation in expressions:

\begin{itemize}
\item
  \verb!* q! means: memory cell whose adress is \verb!q!. (Note that
  \verb!q! must be an address)
\item
  \verb!& a! means: address of \verb!a!. (Note that \verb!a! must be
  an l-value)
\item
  \verb!p + 1! means: the address following the the address \verb!p!. (\verb!p! must be an address)
\end{itemize}

\begin{itemize}
\item Adresses and L-values\hfill{\textbf{[4 points]}}

Which of the following these expressions are l-values? Which
expressions are pointers? (\verb!a!, \verb!b! denote integers
variables; \verb!p!, \verb!q! denote variables containing pointers
to integers.)

Reproduce the following table and replace the question marks with
``yes'' or ``no'' appropriately.

\begin{center}
\begin{tabular}{>{\PBS\raggedright\hspace{0pt}}p{0.15\columnwidth}>{\PBS\raggedright\hspace{0pt}}p{0.19\columnwidth}>{\PBS\raggedright\hspace{0pt}}p{0.10\columnwidth}}
expression
 & l-value
 & pointer
\\
\hline
a
 & ?
 & ?
\\
p
 & ?
 & ?
\\
p + 1
 & ?
 & ?
\\
\& p
 & ?
 & ?
\\
* (p + 1)
 & ?
 & ?
\\
(\& a) + 1
 & ?
 & ?
\\
\end{tabular}
\end{center}


\newpage

\item Parameter passing

Consider the following program.

\begin{verbatim}
f (a, b : integers passed by value-result) {    
    a := b
    b := b - 3;
    return;
}

i: integer
x: array of integers
i := 1;
x[0] := 3;
x[1] := 4;
f(i,x[0]);
print (i + x[0] + x[1]);
\end{verbatim}

What is printed? \hfill{\textbf{[2 points]}}

  
Translation to call-by-reference. \hfill{\textbf{[6 points]}}

Translate the function \verb!f! \emph{and} its call to a language that
does not support call by value-result, but only call by copy.
Do so by using temporary variables and pointers (address variables).  You are not
allowed to change anything else. In particular, the ``algorithm'' and
the declatations of \verb!x! and \verb!i! must remain the same. The
value printed must remain the same.  Assume copy semantics for
assignment and early l-value (memory location) computation.


\end{itemize}


\newpage
\section{Object-Oriented Programming [12 points]}

\begin{description}

\item[Subtyping]
  For this question, we use the notation $A :< B$ to mean $A$ is a
  subtype of $B$.

\begin{enumerate}
\item State the substitution principle of Liskov.  \hfill{\textbf{[2
      points]}}

\item Assume the following subtyping statements:
$Cat :< Animal$, $Dog :< Animal$, $Terrier :< Dog$

Given the above, which of the following subtyping statements are
sound with respect to the substitution principle? (We have seen
sound subtyping rules in the course, apply them to the following
examples) Reproduce the following table and replace the question marks with
``yes'' or ``no'' appropriately.
\hfill{\textbf{[4 points]}}

\newcommand{\mutable}{\mathbf{mutable}}
\newcommand{\constant}{\mathbf{constant}}

\begin{center}
\begin{tabular}{cc}
sound?
 & statement
\\
\hline
?
 & $Cat :< Dog$
\\
?
 & $Terrier :< Animal$
\\
?
 & $Animal → Dog :< Dog → Animal$
\\
?
 & $Dog → Animal :< Animal → Dog$
\\
?
 & $Animal → Animal :< Dog → Dog$
\\
?
 & $Terrier → Terrier :< Dog → Dog$
\\
?
 & $\{ f : Dog \} :< \{ \}$
\\
?
 & $\{ \} :< \{ f : Dog \}$
\\
\end{tabular}
\end{center}

The notation $\{ fieldName : Type \}$ stands for a record with one
immutable field, and $\{ \}$ stands for a record with no field.


\end{enumerate}

\item[Algebraic specification]

Consider the following signature for the sort \verb!S!.

\begin{verbatim}
yes : S
no : S
maybe : S
and : S × S → S
or : S × S → S
not : S → S
\end{verbatim}

Write \emph{at most four} axioms that, together, make all the
following statements true \emph{at the same time}. \\ \hfill{\textbf{[6 points]}}

\begin{itemize}
\item
  and(maybe,maybe) = maybe
\item
  and (yes,no) = or(maybe,no)
\item
  and (maybe,yes) = not(maybe)
\item
  and (no,yes) = or(no,no)
\item
  or(maybe,yes) = or(yes,maybe)
\end{itemize}
\end{description}


\newpage


\end{document}
