\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{listings}
\lstnewenvironment{code}{\lstset{language=Haskell,basicstyle=\small\ttfamily}}{}

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}

\usepackage{pgf, tikz}
\usetikzlibrary{shapes}
\usetikzlibrary{shapes.multipart}
\usetikzlibrary{arrows}

\lstdefinelanguage{FSharp}
      {morekeywords={let, new, match, inherit, abstract, with, rec,
          open, module, namespace, type, of, member, and, for, in, do,
          begin, end, fun, function, try, mutable, if, then, else,
          class, interface, end},
    keywordstyle=\color{blue},
    basicstyle = \small,
    sensitive=false,
    morecomment=[l][\color{green!50!black!80}]{///},
    morecomment=[l][\color{green!50!black!80}]{//},
    morecomment=[s][\color{green!50!black!80}]{{(*}{*)}},
    morestring=[b]",
    stringstyle=\color{red},
    columns=fullflexible
    }


\lstdefinelanguage{Smalltalk}{
  basicstyle=\ttfamily,
  keywordstyle=\bfseries,
  morekeywords={self,super,true,false,nil,thisContext}, % This is overkill
  morestring=[d]',
  morecomment=[s]{"}{"},
  alsoletter={\#:},
  escapechar={!},
}[keywords,comments,strings]


\newcommand{\Blue}[1]{\color{blue}#1\color{black}\xspace}


\usepackage{array}
% This is needed because raggedright in table elements redefines \\:
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\let\PBS=\PreserveBackslash

%%%%%%%%%%%%%% 
\newcommand{\solution}[1] {}
%\newcommand{\solution}[1] {\textbf{Solutions:}\\ #1}

%\newcommand{\comment}[1]{}
\newcommand{\comment}[1]{\marginpar{#1}}


\begin{document}
\newcommand{\examtime}{14:00, March 9th, 2012}
\newcommand{\points}[1]{\marginpar{\bf #1 points}}
\noindent
\begin{tabular}{lr}
CHALMERS TEKNISKA H\"OGSKOLA &\examtime{}.\\
Dept. of Computer Science and Engineering & Programming Paradigms\\
Jean-Philippe Bernardy                 & DAT121 / DIT331(GU) \\
\end{tabular}

\vspace{2.5cm} \noindent
\begin{center} {\LARGE
Exam in Programming Paradigms}
\end{center}

\vspace{1.5cm}

\noindent
\examtime{}.\\
\begin{tabular}{ll}
\textbf{Examiner:} &  John Hughes  \\
\textbf{Lecturer:} & Jean-Philippe Bernardy
\end{tabular}
\vspace{1cm}

\noindent
Permitted aids: Pen and paper. 
% English-Swedish or English-other language dictionary.

There are 6 questions: each worth 10 points. The total sum is 60
points.  

Some questions come with \emph{remarks}: you must take those
into account.  Some questions come with \emph{hints}: you may ignore
those.

You will be asked to write programs in various paradigms in the
following. Choose the language appropriately in each case, and
indicate which you choose at the beginning of your answer.

\begin{tabular}[p]{cc}
  Paradigm & Acceptable language \\ \hline
  Imperative   & C or (an OO language where you refrain to use Objects) \\
  Object oriented & C++ or Java \\
  Functional & Haskell, ML \\
  Concurrent & Erlang, Concurrent-Haskell \\
  Logic & Prolog, Curry 
\end{tabular}

You may also use pseudo-code ressembling an actual language in the
relevant list. In that case, make sure your code can only be
interpreted in the way you intent (the responsibility lies with
you). In particular, in the case of functional/logic languages,
omitting parentheses is NOT acceptable: \texttt{a b c} is not
acceptable pseudo-code for \texttt{a (b c)}.

  
\textbf{Chalmers:}
24 points is required to pass (grade 3), 36 points is required for
grade 4, and 48 points is required for grade 5. 

\textbf{GU:} 
24 points is required to pass (grade G) and 42 points is
required for grade VG.


\newpage


\section{Explicit gotos}

Consider the following optimised code for array copy:
\begin{verbatim}
  short *to, *from;
  int count;
  ...
  {
     /* pre: count > 0 */
     int n = (count + 7) / 8;
     switch(count % 8){
       case 0:	do{     *to++ = *from++;
       case 7:	        *to++ = *from++;
       case 6:	        *to++ = *from++;
       case 5:	        *to++ = *from++;
       case 4:	        *to++ = *from++;
       case 3:	        *to++ = *from++;
       case 2:	        *to++ = *from++;
       case 1:	        *to++ = *from++;
                } while (--n > 0);
     }
  }
\end{verbatim}

Translate the above snippet to equivalent code that does not use
\texttt{switch} nor \texttt{while}: use explicit gotos instead.
Write your answer in a C-like language.

Remark: you must retain the optimisation. Do not test the loop
condition after each copy of a byte.

Note: You can get partial credit for removing either the switch or the
while.

\newpage
\section{Objects from records}

Consider the following C++ code, which is an encoding of a class
hierarchy.
\begin{verbatim}
struct C {
   float x;
   float y;
};

struct B {
   void *n(B*, C*);
   int f;
   void *m(B*, B*);
   int g;
};

struct A {
   void *m(A*, B*);
   int f;
};

void B_n(B* p, C* q) {
  q->x += p->f;
  q->y += p->g;
}

void B_m (B* p, B* q) {
  q->f += p->f;
  q->g += p->g;
}

void A_m(A* p, B* q) {
  q->f += p->f;
}
\end{verbatim}

Quick refresher:
\begin{itemize}
\item If \texttt{A} is a type, then \texttt{A*} is the type of
  pointers to \texttt{A}.
\item \texttt{void *f(A,B,...)} stands for a pointer to a function
  \texttt{f} taking arguments \texttt{A},\texttt{B},...
\item \texttt{p->f} is used to access a field \texttt{f} in a
  structure pointed by \texttt{p}.
\end{itemize}

Write the original class hierarchy in a C++-like or Java-like language.


\newpage
\section{Higher-order abstractions}

Consider the following Haskell code:
\begin{verbatim}
concat [] = []
concat (x:xs) = x ++ concat xs
\end{verbatim}


% data Tree = B Tree Tree | Leaf Int
% 
% reverse (Leaf x) = (Leaf (0-x))
% reverse (B x y) = B (reverse y) (reverse x)


Express the same function in terms of the higher-order functions
provided below.  Remark: you may \emph{not} use recursion directly in
your solution.

\begin{verbatim}
fold k f [] = k
fold k f (x:xs) = f x (fold k f xs)
\end{verbatim}

% foldT k f (Leaf x) = k x
% foldT k f (Bin x y) = f (foldT k f x) (foldT k f y)


\section{Closures}

Consider the Haskell program:
\begin{verbatim}
f m grades bonuses = zipWith (\x y -> x + y) bonuses (filter (\ x -> x < m) grades)

zipWith f [] [] = []
zipWith f (x:xs) (y:xs) = f x y : zipWith f xs ys

filter p [] = []
filter p (x:xs) = if p x
                  then x : filter p xs
                  else filter p xs
\end{verbatim}

Translate the code to use explicit closures. 

Remarks:
\begin{itemize}
\item All higher-order functions must be removed.
\item You cannnot \emph{specialize} any higher-order function
  (eg. don't make a special version of \texttt{zipWith} which does
  only addition).  Use closures.
\end{itemize}

\newpage
\section{Laziness}

In Haskell you can write a matrix comprehension like so:
\begin{verbatim}
matrix = array bounds [((x,y),valueAtIndex x y) | (x,y) <- range bounds]
  where bounds = ((low_x,low_y),(high_x,high_y))
\end{verbatim}

You can then access a cell $x,y$ in the array using \texttt{matrix!(x,y)}.

For example, in this matrix of $100$ elements the value at index $(x,y)$ contains $x ×
y$: \texttt{example!(x,y) == x * y}.
\begin{verbatim}
example = array bounds [((x,y),x * y) | (x,y) <- range bounds]
  where bounds = ((1,1),(10,10))
\end{verbatim}


Consider the following algorithm to compute the Levenshtein distance between two strings:
{\small
\begin{verbatim}
int LevenshteinDistance(char s[1..m], char t[1..n])
{
  // for all i and j, d[i,j] will hold the Levenshtein distance between
  // the first i characters of s and the first j characters of t;
  // note that d has (m+1)x(n+1) values
  declare int d[0..m, 0..n]

  for i from 0 to m
    d[i, 0] := i // the distance of any first string to an empty second string
  for j from 0 to n
    d[0, j] := j // the distance of any second string to an empty first string

  for j from 1 to n {
    for i from 1 to m {
      if s[i] = t[j] then
        d[i, j] := d[i-1, j-1]       // no operation required
      else
        d[i, j] := minimum
                   (
                     d[i-1, j] + 1,  // a deletion
                     d[i, j-1] + 1,  // an insertion
                     d[i-1, j-1] + 1 // a substitution
                   )
    }
  }
  return d[m,n]
}
\end{verbatim}
}

Using \emph{a single comprehension}, construct a matrix \texttt{d},
such that \texttt{d!(i,j)} is the Levenshtein distance between the
first \texttt{i} characters of \texttt{s} and the first \texttt{j}
characters of \texttt{t}, assuming
\begin{verbatim}
s :: String
t :: String
\end{verbatim}

Remark: The distance between any two substrings must not be computed
twice. Use laziness!

Hints:
\begin{itemize}
\item You may also assume a predefined function \texttt{minimum}:
\begin{verbatim}
   minimum :: [Int] -> Int
\end{verbatim}
\item You can access the $\mbox{\texttt{i}}^{\mbox{th}}$ character of
  string \texttt{s} with the expression \texttt{s!!(i-1)}.
\end{itemize}

\section{Functions to relations}

Consider the function
\begin{verbatim}
fib :: Int -> Int
fib 0 = 0
fib 1 = 1
fib n = fib (n-1) + fib (n-2)
\end{verbatim}

Translate the function \texttt{fib} to a relation \texttt{fibo}, such that
\begin{verbatim}
       fibo x y      is equivalent to      fib x == y
\end{verbatim}

Remarks:
\begin{itemize}
\item You may not use the \emph{functions} \texttt{fib} nor \texttt{(+)} in the
  translated version.
\item Use a relation \texttt{plus} which coincides with addition:
  \begin{verbatim}
       plus x y z     is equivalent to      x + y == z
\end{verbatim}

\begin{verbatim}

\end{verbatim}
\end{itemize}

Hint: you may want to start by writing the types of the
\emph{relations} \texttt{fibo} and \texttt{plus}.

You may write your answer in Curry syntax or Prolog syntax. 

\end{document}
